\bf Interactive Visualization of Hierarchically Structured Data

### 2017-07-12 treelapse.tex

\citet{dethlefsen2008pervasive} investigated the effect of antibiotics on
bacterial community composition from an ecological perspective. The study tracks
the microbiome of three patients across ten months, with two five-day antibiotic
time courses separated by 6 months. Discerning the variation in resilience
across bacteria is important, considering the the role of bacteria in health and
not just disease.

@@

\caption{Here we display the primary timebox tree view of the antibiotics data
  set from \citep{dethlefsen2008}, annotated with the main components of the
  visualization. The tree at the top is a taxonomic tree of all the bacteria
  contained within the sample, and it is visually linked to the time series at
  the bottom: each node in the tree corresponds to a path among the time series.
  The selection brush is used to focus attention on the time series that go
  through it -- these are highlighted in blue -- and other brushes can be added
  using a button not displayed here. The pan-zoom widget at the top right is
  used to update the scales of the main time series display so that only
  particular time windows and $y$-axis ranges are visible. This view is the
  basis for all the timebox tree and treebox displays that appear
  below. \label{antibioticoverview}}
\end{figure}

@@

\caption{Introducing a second box into the timebox display identifies the
  Bacteroidetes as a taxon only midly impacted by antibiotics. The layout is
  identical to Figure 1, except two small brushes are placed over the time
  series between 10 and 20 days, and now only those time series and
  corresponding nodes in the tree are highlighted in blue. Further, the user has
  hovered over the top blue node in the tree, revealing the taxonomic identity
  of these series. Hence, brushing the time series and linking with the tree can
  be used to discover and characterize notable variation within the
  data.\label{fig:antibioticbacteroidetes}}
\end{figure}

@@

\caption{Zooming into the second antibiotic timecourse and highlighting the
  Lachnospiraceae and Ruminococcus, we see that the Ruminoccocus took more time
  to recover to pre-treatment levels. Here, the red lines and nodes are those
  that match the text search provided by the user in a search box just below the
  figure (not displayed here). Hovering the mouse over these lines provides
  their identities -- the top red line is Lachnospiraceae, and the bottom red
  line is Ruminoccocus. Note that the brush in the treebox display is located
  over the tree, rather than over the time series. In particular, the search box
  and interactive brushing can be combined to interrogate hypotheses of apriori
  interest. \label{fig:antibioticfirmicutes}}
\end{figure}

@@

{\centering \includegraphics[width=300px]{figure/preterm_preterm}

\caption{Samples with high levels of phyla other than Firmicutes appear to be
  related to preterm births. Here, we again display the preterm birth data with
  a DOI sankey, but instead of grouping samples according to statistically
  generated CSTs, we directly assign samples to preterm vs. not preterm
  according to whether the mother eventually had a preterm birth. These new
  states are visible in the updated legend. Through interactivity, it becomes
  possible to develop meaningful visual summaries even before calculating formal
  statistical ones.}
}\label{fig:pretermpreterm}
\end{figure}

@@

Alternatively, we can study the trajectories of home prices among neighborhoods,
conditional on their being middle-income before the recession. We generate the
sequence of views in Figures \ref{fig:zillowmiddlepre},
\ref{fig:zillowmiddleup}, and \ref{fig:zillowmiddledown} to this end. The first
of these figures isolates neighborhoods with middle incomes before the
recession, using a single timebox. Since there appears to be a divergence in
trajectories after the recession, we introduce a second post-recession timebox,
dragging it over series with higher and lower incomes during this second time
period. This is the content of Figures \ref{fig:zillowmiddleup} and
\ref{fig:zillowmiddledown}. Inspecting the highlighted tree nodes associated
with these series, we find that most of the middle-income series that increased
after the recession are associated with middle-income neighborhoods within the
coastal Southern California counties. In constrast, those middle income series
that saw decreases were mostly located in Central California and Oakland.

@@

\caption{California home prices at the city level, between 1996 and 2016. The
  effect of the 2008 recession is clear, and we have hovered over the San
  Bernardino series, to display the identity of one of the cities most strongly
  affected by the recession.}\label{fig:zillowcities}
\end{figure}

@@

\caption{Neighborhoods with mid-range home prices before the recession are
  selected here. Note that the collection of series seems to widen after 2008 --
  we are interested in whether there are reliable predictors of these alternate
  trajectories, given their similar starting points.}\label{fig:zillowmiddlepre}
\end{figure}

@@

\caption{Among those neighborhoods with mid-range prices before the recession,
  we have selected those that recovered more rapidly. These appear to be located
  mainly in Los Angeles and San Diego counties.}\label{fig:zillowmiddleup}
\end{figure}

@@

\caption{Rather than selecting mid-range series that recovered quickly, we can
  isolate those whose prices remained depressed after the recession. These seem
  mostly to be located in Central California and the East San Francisco Bay
  Area.}\label{fig:zillowmiddledown}
\end{figure}

@@

\caption{To study the range in home prices within San Francisco County, we can
  select the associated nodes using a treebox.}\label{fig:zillowsf}
\end{figure}

@@

\caption{The two peaks at rush hour distinguish weekday series from the
  rest.}\label{fig:working}
\end{figure}

@@

{\centering \includegraphics[width=350px]{figure/weekend}

\caption{Unlike weekday demand, weekend demand is unimodal. The few weekday
  series with unimodal series seem to be associated with holidays. This is the
  case for New Years' Eve, which is currently hovered over in the
  tree.}\label{fig:weekend}
\end{figure}

{\centering \includegraphics[width=350px]{figure/weekday_2011}

\caption{Weeday demand appears larger in 2012 than 2011 -- compare with the next
  figure.}\label{fig:weekday2011}
\end{figure}

@@

\caption{Weekday demand increased in 2012 -- compare with the previous
  figure.}\label{fig:weekday2012}
\end{figure}

@@

\caption{The samples with highest night demand tend to fall on warm
  weekends.}\label{fig:warmweekend}
\end{figure}

@@

\caption{We can interpret the regression tree using an interactive DOI
  representation.}\label{fig:bikesankey}
\end{figure}

This interactive representation of regression trees is potentially more
useful on larger trees that cannot be easily parsed in a single view; in
this sense the bikesharing tree is relatively simple. In our ideal data
analysis workflow, we imagine the analyst applying interactive
visualization and modeling techniques in an iterative, nonlinear
fashion.

@@

\caption{An application to the Global Patterns demonstrates how linking in
  treelapse can be applied to combine hierarchical clustering and parallel
  coordinates views.}\label{fig:gptimebox}
\end{figure}

@@

\caption{Viewing a tree of $p$-values across different methods highlights two
  subtrees with strong associations with mouse age, across several testing
  procedures.}\label{fig:structssi}
\end{figure}

Further, selecting series with strong association between abundance and age, two
major subtrees are brought to the forefront. Separately querying the the
taxonomic identities of these bacteria reveals that they are two subgroups of
Clostridia, which is consistent with the analysis of
\citep{callahan2016bioconductor}. More than this specific analysis outcome, this
view demonstrates that interactive visual inspection of results from
confirmatory analysis provides deeper insight than the standard practice of
printing tables of (adjusted or unadjusted) $p$-values: the relationship between
significant nodes is only clear upon visualization on the tree.

>>>>

\citet{dethlefsen2008pervasive} investigated the effect of antibiotics on
bacterial community composition from an ecological perspective. The study tracks
the microbiome of three patients across ten months, with two five-day antibiotic
time courses separated by 6 months. Discerning the variation in resilience
across bacteria is important, considering the role of bacteria in health and
not just disease.

@@

\caption{Here we display the primary timebox tree view of the antibiotics data
  set from \citep{dethlefsen2008pervasive}, annotated with the main components of the
  visualization. The tree at the top is a taxonomic tree of all the bacteria
  contained within the sample, and it is visually linked to the time series at
  the bottom: each node in the tree corresponds to a path among the time series.
  The selection brush is used to focus attention on the time series that go
  through it -- these are highlighted in blue -- and other brushes can be added
  using a button not displayed here. The pan-zoom widget at the top right is
  used to update the scales of the main time series display so that only
  particular time windows and $y$-axis ranges are visible. This view is the
  basis for all the timebox tree and treebox displays that appear
  below. \label{fig:antibioticoverview}}
\end{figure}

@@

\caption{Introducing a second box into the timebox display identifies the
  Bacteroidetes as a taxon only mildly impacted by antibiotics. The layout is
  identical to Figure 1, except two small brushes are placed over the time
  series between 10 and 20 days, and now only those time series and
  corresponding nodes in the tree are highlighted in blue. Further, the user has
  hovered over the top blue node in the tree, revealing the taxonomic identity
  of these series. Hence, brushing the time series and linking with the tree can
  be used to discover and characterize notable variation within the
  data.\label{fig:antibioticbacteroidetes}}
\end{figure}

@@

\caption{Zooming into the second antibiotic time course and highlighting the
  Lachnospiraceae and Ruminococcus, we see that the Ruminoccocus took more time
  to recover to pre-treatment levels. Here, the red lines and nodes are those
  that match the text search provided by the user in a search box just below the
  figure (not displayed here). Hovering the mouse over these lines provides
  their identities -- the top red line is Lachnospiraceae, and the bottom red
  line is Ruminoccocus. Note that the brush in the treebox display is located
  over the tree, rather than over the time series. In particular, the search box
  and interactive brushing can be combined to interrogate hypotheses of a priori
  interest. \label{fig:antibioticfirmicutes}}
\end{figure}

@@

{\centering \includegraphics[width=300px]{figure/preterm_preterm}}

\caption{Samples with high levels of phyla other than Firmicutes appear to be
  related to preterm births. Here, we again display the preterm birth data with
  a DOI sankey, but instead of grouping samples according to statistically
  generated CSTs, we directly assign samples to preterm vs. not preterm
  according to whether the mother eventually had a preterm birth. These new
  states are visible in the updated legend. Through interactivity, it becomes
  possible to develop meaningful visual summaries even before calculating formal
  statistical ones.}
\label{fig:pretermpreterm}
\end{figure}

@@

Alternatively, we can study the trajectories of home prices among neighborhoods,
conditional on their being middle-income before the recession. We generate the
sequence of views in Figures \ref{fig:zillowmiddlepre},
\ref{fig:zillowmiddleup}, and \ref{fig:zillowmiddledown} to this end. The first
of these figures isolates neighborhoods with middle incomes before the
recession, using a single timebox. Since there appears to be a divergence in
trajectories after the recession, we introduce a second post-recession timebox,
dragging it over series with higher and lower incomes during this second time
period. This is the content of Figures \ref{fig:zillowmiddleup} and
\ref{fig:zillowmiddledown}. Inspecting the highlighted tree nodes associated
with these series, we find that most of the middle-income series that increased
after the recession are associated with middle-income neighborhoods within the
coastal Southern California counties. In contrast, those middle income series
that saw decreases were mostly located in Central California and Oakland.

@@

\caption{California home prices at the city level, between 1996 and 2016. Here,
  we provide a treebox display, as in Figure \ref{fig:antibioticfirmicutes}, but
  of the California home prices data. In this case, the tree corresponds to a
  geographic hierarchy, with regions at the top and neighborhoods at the bottom, while
  time series correspond to home prices collected weekly between 1996 and 2016.
  A brush is drawn over all nodes at the city level. The effect of the 2008
  recession is clear, and we have hovered over the San Bernardino series, to
  display the identity of one of the cities most strongly affected by the
  recession. This provides an example of a dataset with meaningful (spatial)
  hierarchical structure that might not initially be studied using trees.
\label{fig:zillowcities}}
\end{figure}

@@

\caption{Neighborhoods with mid-range home prices before the recession are
  brushed here, using a timebox tree version of Figure \ref{fig:zillowcities}.
  The associated tree nodes are highlighted at the top. Note that the collection
  of series seems to widen after 2008 -- we are interested in whether there are
  reliable predictors of these alternate trajectories, given their similar
  starting points. This serves as a baseline with which to compare Figure
  \ref{fig:zillowmiddleup} -- these views are easy to transition between
  interactively.}\label{fig:zillowmiddlepre}
\end{figure}

@@

\caption{Among those neighborhoods with mid-range prices before the recession,
  displayed in Figure \ref{fig:zillowmiddlepre}, we have selected those that
  recovered more rapidly by introducing a brush at the top right of the time
  series panel. By hovering a brush over a collection of tree nodes that all
  seem to be highlighted, we infer that the associated neighborhoods are located
  mainly in Los Angeles and San Diego counties. This follow-up view is
  interesting to contrast with Figure \ref{fig:zillowmiddledown}.
}\label{fig:zillowmiddleup}
\end{figure}

@@

\caption{In contrast to Figure \ref{fig:zillowmiddleup}, we can follow-up the
  selection in Figure \ref{fig:zillowmiddlepre} by isolating those neighborhoods
  where prices remained depressed after the recession. This is accomplished by
  moving the brush on the right down towards lower prices. Hovering over the
  associated highlighted nodes in the tree to reveal the associated locations,
  we see that most of these series correspond to neighborhoods in Central
  California and the East San Francisco Bay Area. The ability to sketch the
  overall shapes of series using brushes and interpret the associated selections
  using a tree simplifies what might otherwise be complex
  comparisons.}\label{fig:zillowmiddledown}
\end{figure}

@@

\caption{In contrast to Figures \ref{fig:zillowmiddlepre} to
  \ref{fig:zillowmiddledown}, we can study the variation in series associated
  with a subset of the tree by using treeboxes. To study the range in home
  prices within San Francisco County, we can brush over the associated nodes in
  the tree. The red circle indicates that the user has searched for ``San
  Francisco,'' which guides the user to the appropriate subtree. The mouse is
  hovering over one of the more expensive neighborhoods. Hence, though the
  timebox tree and treebox views are similar, they are directed towards
  different types of visual comparisons.}\label{fig:zillowsf}
\end{figure}

@@

\caption{The two peaks at rush hour distinguish weekday series from the rest.
  through the timebox tree view. The display is the same type of timebox tree
  view introduced in Figure \ref{fig:antibioticoverview}, but applied to the
  bikesharing data, where the time axis represents the time of day and the
  $y$-axis gives bikesharing demand. Each series is the bikesharing demand for a
  single day, over the course of two years. The tree now corresponds to the
  regression tree generated by predicting demand at 8am using supplementary
  data. Two brushes are introduced to highlight the double peaks corresponding
  to rush hours on weekdays. We see that although hierarchical structure was not
  present immediately in the bikesharing data, it is useful to introduce and
  interpret such structure by combining regression and visualization
  methodology.
}\label{fig:working}
\end{figure}

@@

{\centering \includegraphics[width=350px]{figure/weekend}}

\caption{By adjusting the two brushes in Figure \ref{fig:working}, we see that
  unlike weekday demand, weekend demand is unimodal. The few weekday series with
  unimodal series seem to be associated with holidays. This is the case for New
  Years' Eve, which is currently hovered over in the tree. The ease of
  transitioning from Figure \ref{fig:working} to this display indicates the
  importances of brushing in interaction.}\label{fig:weekend}
\end{figure}

\begin{figure}
{\centering \includegraphics[width=350px]{figure/weekday_2011}}

\caption{Weekday demand appears larger in 2012 than 2011 -- compare with Figure 
  \ref{fig:weekday2012}. Here, a brushes are introduced over the tree to see the
  series associated with a particular split point. The red nodes are the results
  of searches over the two nodes that are children of this split point. The fact
  that the \texttt{yr>=0.5} selected line is larger than the \texttt{yr<0.5}
  line means that demand was larger in 2012. In combination with searching and
  treeboxes, it is possible to interpret more subtle split points in the decision
  tree.}
\label{fig:weekday2011}
\end{figure}

@@

\caption{Weekday demand increased in 2012 -- compare with 
  Figure \ref{fig:weekday2011}. The search terms are the same as in that figure,
  but the subtree associated with the 2012 split point is highlighted, using the
  union of two boxes. Note that unlike timebox trees, which highlighted series
  lying through the intersection of brushes, treeboxes highlight series within the
  union of brushes.}\label{fig:weekday2012}
\end{figure}

@@

\caption{The samples with highest night demand tend to fall on warm
  weekends. Here, the pan-zoom widget has been used to adjust both time and
  demand axes to narrow on a specific window of interest. Brushes are drawn over
  the larger among these series, and the corresponding tree nodes are located
  close to one another, in the part of the tree corresponding to the warm / cold
  average temperature split. More generally, panning and zooming allows
  navigation between focus and context.}\label{fig:warmweekend}
\end{figure}

@@

\caption{So far, we have focused on the timebox tree and treebox representations
  of the bikesharing data -- a complementary view is provided by the DOI sankey.
  Here, the tree is the result of the regression tree procedure, while the
  colors represent particular quantiles of 8am demand. This allows the
  determination of which split descendants are associated with low or high
  demand.}\label{fig:bikesankey}
\end{figure}

This interactive representation of regression trees is potentially more
useful on larger trees that cannot be easily parsed in a single view; in
this sense the bikesharing tree is relatively simple. In our ideal data
analysis workflow, we imagine the analyst applying interactive
visualization and modeling techniques in an iterative, nonlinear
fashion, in the spirit of \citep{de2003visual}.

@@

\caption{An application to the Global Patterns demonstrates how linking in
  treelapse can be applied to combine hierarchical clustering and parallel
  coordinates views. Each panel represents a different subtree cluster within
  this data set, as indicated by the different locations for the tree brushes.
  The paths in the lower halves of each display represent the average value
  across different bacteria rather than timepoints, as in all previous figures.
  Though the samples originally do not include any hierarchical structure,
  hierarchical clustering provides such a structure which can then be
  interpreted using treeboxes. }\label{fig:gptimebox}
\end{figure}

@@

\caption{Viewing a tree of $p$-values across different methods highlights two
  subtrees with strong associations with mouse age, across several testing
  procedures. The tree represents the taxonomy of bacteria, and the series
  provides the negative log $p$-values associated with nodes as computed by
  different tests, listed along the $x$-axis as in parallel coordinates. By
  selecting series with larger values for a test, we see the associated subtrees
  of significant $p$-values. Hence, hierarchical views can be useful even in the
  confirmatory testing settings which typically study results from individual
  tests in isolation from each other.\label{fig:structssi}}
\end{figure}

Further, selecting series with strong association between abundance and age, two
major subtrees are brought to the forefront. Separately querying the
taxonomic identities of these bacteria reveals that they are two subgroups of
Clostridia, which is consistent with the analysis of
\citep{callahan2016bioconductor}. More than this specific analysis outcome, this
view demonstrates that interactive visual inspection of results from
confirmatory analysis provides deeper insight than the standard practice of
printing tables of (adjusted or unadjusted) $p$-values: the relationship between
significant nodes is only clear upon visualization on the tree.
